\chapter{Implementation}

\section{Project Organisation}

The project is organised as a modular MATLAB/Simulink workspace. Aircraft data, environment constants, aerodynamic coefficients and reference cases are stored separately from plant dynamics and controller code. This structure supports incremental validation and makes it easier to replace simplified modules with more accurate versions later in the project.

\begin{table}[H]
    \centering
    \caption{Main project directories.}
    \begin{tabular}{ll}
        \toprule
        Directory & Purpose \\
        \midrule
        \texttt{data} & aircraft, atmosphere, aerodynamic and reference-case data \\
        \texttt{plant} & nonlinear six-degree-of-freedom plant implementation \\
        \texttt{nmpc} & predictive controller models and controller-generation scripts \\
        \texttt{guidance} & mode logic and trajectory-reference generation \\
        \texttt{validation} & trim, dynamic-response and autopilot validation scripts \\
        \texttt{models} & Simulink model files \\
        \texttt{tests} & feature-oriented MATLAB checks \\
        \texttt{tools} & DATCOM extraction, validation and reporting helpers \\
        \texttt{docs} & documentation and thesis sources \\
        \bottomrule
    \end{tabular}
\end{table}

\section{Data Modules}

The current implementation contains the following data modules:
\begin{itemize}
    \item \texttt{b777\_constants.m}, which defines project-wide constants such as standard gravity,
    \item \texttt{b777\_geometry.m}, which defines B777-300ER-like geometry and reference dimensions,
    \item \texttt{b777\_mass\_inertia.m}, which defines mass envelope values and first-cut inertia estimates,
    \item \texttt{b777\_atmosphere\_isa.m}, which provides a simple ISA atmosphere model,
    \item \texttt{b777\_engine\_model.m}, which defines the first B777-like propulsion seed and thrust-lapse interface,
    \item \texttt{b777\_aero\_database.m}, which defines the staged aerodynamic database interface and source inventory,
    \item \texttt{b777\_aero\_clean\_crm\_seed.m}, which exposes the NASA CRM clean Mach-alpha grid,
    \item \texttt{b777\_aero\_read\_crm\_clean\_grid\_csv.m}, which reads the curated NASA CRM clean grid,
    \item \texttt{b777\_aero\_read\_crm\_force\_moment\_csv.m}, which reads NASA CRM force and moment CSV files,
    \item \texttt{b777\_aero\_read\_derivative\_csv.m}, which reads source-oriented derivative tables,
    \item \texttt{b777\_aero\_read\_mach\_derivative\_csv.m}, which reads source-oriented Mach-dependent derivative candidate tables,
    \item \texttt{b777\_aero\_read\_high\_lift\_csv.m}, which reads source-oriented high-lift increment tables,
    \item \texttt{b777\_aero\_derivative\_seed.m}, which exposes the DATCOM-ready static, control and damping residual derivatives,
    \item \texttt{b777\_aero\_high\_lift\_seed.m}, which exposes the CRM-HL/DATCOM-ready clean, approach and landing configuration increments,
    \item \texttt{b777\_aero\_simple.m}, which defines a first-cut aerodynamic coefficient model,
    \item \texttt{b777\_configs.m}, which defines aircraft configurations,
    \item \texttt{reference\_cases.m}, which defines representative flight conditions for trim and validation,
    \item \texttt{conventions.m}, which records units, frames, states and input conventions.
\end{itemize}

The data modules use explicit field names containing units where practical. This convention is intentionally verbose, but it reduces ambiguity and supports code review. For example, mass values are stored with the suffix \texttt{\_kg}, while speeds use \texttt{\_mps}.

\subsection{Aerodynamic Database Build Workflow}

The aerodynamic database was implemented as a sequence of reproducible data
preparation steps. The workflow is intentionally conservative: each generated
candidate table carries a source identifier and remains separate from the
active plant table until validation and review criteria are satisfied.

The practical build sequence is:
\begin{enumerate}
    \item define the B777-like reference geometry and mass properties used for
    aerodynamic scaling;
    \item read the curated NASA CRM clean Mach-alpha grid and expose it as the
    active clean-cruise longitudinal baseline;
    \item preserve the analytical seed model as a complete fallback model for
    states and coefficient groups not covered by validated data;
    \item generate and run Digital DATCOM clean static and damping decks at the
    selected Mach stations;
    \item parse the DATCOM output, convert printed derivatives to project
    units and consolidate the per-Mach outputs into a candidate derivative
    table;
    \item generate and run elevator and aileron control-effectiveness decks,
    then consolidate direct control-derivative candidates;
    \item derive low-confidence gap-fill candidates for active-interface terms
    not printed directly by the current DATCOM workflow;
    \item solve clean-cruise longitudinal trim cases for the candidate
    aerodynamic path;
    \item linearise a restricted longitudinal model at the same trim points and
    check the phugoid and short-period modes;
    \item evaluate promotion gates before any candidate derivative replaces the
    active seed table.
\end{enumerate}

The repository contains the scripts required to reproduce those steps. The
script \path{tools/run_digital_datcom.sh} downloads the public-domain DATCOM
archive from PDAS, compiles the legacy Fortran source locally and runs a
committed B777-like input deck. The extractor
\path{tools/extract_datcom_derivatives.py} parses the static and dynamic
derivative tables, converts derivatives printed per degree to derivatives per
radian and records skipped cells such as \texttt{NDM}.

The wrapper \path{tools/run_datcom_mach_grid.sh} runs the committed clean
single-condition decks at $M=0.30$, $0.50$, $0.60$, $0.70$ and $0.80$.
\path{tools/build_datcom_mach_grid_table.py} consolidates the per-Mach
candidate files into one Mach-dependent table with a grid-level source
identifier and retained per-run source identifiers.

The companion wrapper \path{tools/run_datcom_control_grid.sh} generates and
runs elevator and aileron control-effectiveness decks over the same Mach grid.
Its extractor, \path{tools/extract_datcom_control_derivatives.py}, parses the
DATCOM high-lift/control-device output and produces source-backed candidates
for elevator and aileron derivatives. The helper
\path{tools/build_datcom_control_grid_table.py} consolidates these control
candidates and records which entries are not direct DATCOM printouts.
\path{tools/build_datcom_gap_fill_candidates.py} then derives low-confidence
active-interface candidates for $C_{D_\beta}$, $C_{Y_r}$, rudder derivatives
and lateral control drag from the DATCOM candidate tables and the B777-like
geometry.

The validation scripts form the final part of the workflow.
\path{tools/run_trim_validation.py} solves clean-cruise longitudinal trim cases
for the DATCOM candidate path and writes both a case table and a report.
\path{tools/run_longitudinal_modal_validation.py} linearises a restricted
longitudinal model around the same trim points and checks the phugoid and
short-period modes. Finally,
\path{tools/evaluate_datcom_promotion_gates.py} evaluates whether the direct
and derived candidate tables are allowed to replace the active derivative seed.
The resulting candidate CSV files, extraction reports, promotion-gate reports,
trim-validation artifacts and modal-validation artifacts are stored under
\path{data/aerodynamics/raw/datcom}, \path{data/aerodynamics/validation} and
\path{docs}. Generated DATCOM run folders are treated as reproducible workspace
outputs and are not part of the source repository.

\section{Startup and Path Management}

The file \texttt{startup.m} configures the MATLAB path by adding the project subdirectories required by the current implementation. It avoids relying on the user's current MATLAB path and makes the execution of feature tests repeatable from a clean MATLAB session.

\section{Plant Adapters and Feature Tests}

The plant directory now includes \texttt{b777\_aero\_forces\_moments.m}, which is the first plant-facing aerodynamic load adapter. It completes the atmosphere-dependent quantities required by the coefficient database, computes dynamic pressure, converts wind-axis lift and drag to body-axis force components, scales aerodynamic moments and applies the configured centre-of-gravity shift.

The plant directory also includes \texttt{b777\_engine\_forces\_moments.m}. This
adapter reads the propulsion seed, computes local maximum and idle thrust from
altitude and Mach number, maps throttle commands to commanded thrust, evaluates
the first-order spool derivatives and converts the left and right engine
thrust states into body-axis forces and moments using the engine positions from
the geometry module.

The project includes feature-oriented MATLAB checks under \path{tests/features}. The project-structure test checks the expected top-level folders. The modelling-conventions test checks frame definitions, state-vector sizes and input-vector sizes. The aircraft-data test validates the current geometry, mass, inertia, atmosphere and reference-case data. The aerodynamic-database test checks the staged aerodynamic database interface, source inventory, derivative inventory, CRM clean-grid values, DATCOM-ready residual derivative effects, the reproducible DATCOM Mach-grid, control-grid and derived gap-fill candidate status, trim-validation and modal-validation evidence, CRM-HL/DATCOM-ready high-lift metadata, clean/approach/landing behaviour, aerodynamic force signs and the reference-point-to-CG moment shift. These checks are not substitutes for physical validation, but they catch interface errors early.

The engine-loads feature test checks static thrust consistency with the
geometry module, altitude/Mach thrust lapse, idle-to-maximum throttle mapping,
symmetric force and pitch-moment generation, asymmetric yaw-moment generation
and the signs of the first-order thrust-spool derivatives.

\section{Current Implementation Status}

At the current stage, the data layer, aerodynamic plant adapter and propulsion
load adapter are available. Geometry, mass, inertia, atmosphere and staged
first-cut aerodynamics have been defined, including the clean CRM Mach-alpha
grid, DATCOM-ready residual derivative metadata, source-backed DATCOM
Mach-grid and control-grid derivative candidate run sets, derived gap-fill
candidates, clean-cruise longitudinal trim-validation and modal-validation
artifacts, CRM-HL/DATCOM-ready high-lift increments and body-axis aerodynamic
loads. The propulsion layer now provides an idle-to-maximum throttle map,
altitude/Mach thrust lapse, first-order spool derivatives and body-axis engine
forces and moments. The next implementation step is to add control-surface
actuator dynamics and assemble the nonlinear six-degree-of-freedom equations in
MATLAB/Simulink.
